Investigating Predictors for American Attitudes on Justifiability of Abortion

### 2022-06-21 previous-work/draft/draft.qmd
FINAL REORG

```{r load-data}
#| message: false
abortion_data_full <- read_csv(here::here("data/abortion-attitudes", "wvs-usa-abortion-attitudes-data.csv"))
```

@@

Understanding public attitudes on divisive political issues is an important way for political leaders to mobilize voters and for lawmakers to draft laws that represent their constituents.  While it is simple to poll constituents’ positions on an issue, it can be challenging to assess the complex factors that influence and predict those stances.

Abortion is one such divisive issue in the United States, with strong organization of pro-choice and pro-life groups in states across the country in support of legislation for their respective sides (Ziegler, 2020). However, following the Supreme Court’s 1973 decision in Roe v. Wade, a ruling that has historically protected women's rights to abortion without excessive government restriction in America, the issue has risen in political salience. Both the pro-choice and pro-life movements have gained national prominence, and the two major political parties have polarized around the issue, with the Democratic Party in favor of and the Republican Party against policies legalization and increased access to abortion (Weinberger 2022). Abortion has also increasingly become a key issue that voters consider, with an increasing share of Americans identifying as “single-issue voters” regarding abortion (Brenan 2020).

@@

We will conduct our analysis on data collected about Americans’ abortion attitudes, demographic information, and other ideologies, using an EDA-informed predictive model. The data are a nationally representative sample of the American people; hence we can infer that all the observations are independent and the variables may show a linear relationship. Given that this issue has become highly polarized by political party, we predict that liberal political attitudes and youth will correlate with belief that abortion is more justified.

@@

The response variable, `Justifiability of abortion`, is a numerical measure on a scale of 1 to 10 on the individual person’s attitude toward whether abortion is justifiable or not.  Individuals responded 1 for “abortion is never justified” and 10 for “abortion is always justified.

@@

```{r visualization-summary}
#| warning: false
ggplot(abortion_data, aes(x = aj)) +
  geom_histogram(binwidth = 1, aes(y=..density..)) + 
  geom_density(color = "red") +
  labs(title = "Histogram of abortion attitudes")

summary(abortion_data$aj)
```

@@

```{r eda-year}
#| warning: false
#Year of survey
ggplot(abortion_data, aes(x = year)) +
  geom_histogram() + 
  labs(title = "Histogram of year")
ggplot(abortion_data, aes(group = year, y = aj)) +
  geom_boxplot() 
```

```{r eda-ideology}
#| warning: false
# ideology
ggplot(abortion_data, aes(x = ideology)) +
  geom_histogram() + 
  labs(title = "Histogram of ideology")
ggplot(abortion_data, aes(x = ideology,group = ideology, y = aj)) +
  geom_boxplot()
```

```{r eda-cai}
#| warning: false
# Child autonomy index
ggplot(abortion_data, aes(x = cai)) +
  geom_histogram() + 
  labs(title = "Histogram of Child autonomy index")
ggplot(abortion_data, aes(x = cai, group = cai, y = aj)) +
  geom_boxplot()
```

```{r eda-god}
#| warning: false
# Importance of God
ggplot(abortion_data, aes(x = godimportant)) +
  geom_histogram() + 
  labs(title = "Histogram of how respondent saw God's importance")
ggplot(abortion_data, aes(x = godimportant, group = godimportant, y = aj)) +
  geom_boxplot()
```

```{r eda-respect}
#| warning: false
# Respect for authority
ggplot(abortion_data, aes(x = respectauthority)) +
  geom_histogram() + 
  labs(title = "Histogram of respect for authority")
ggplot(abortion_data, aes(x = respectauthority, group = respectauthority, y = aj)) +
  geom_boxplot()
```

```{r eda-pride}
#| warning: false
# National Pride
ggplot(abortion_data, aes(x = nationalpride)) +
  geom_histogram() + 
  labs(title = "Histogram of National pride")
ggplot(abortion_data, aes(x = nationalpride, group = nationalpride, y = aj)) +
  geom_boxplot()
```

@@

```{r correlations}
#| warning: false
ggpairs(abortion_data, 
  columns = c("year", "ideology", "cai", "godimportant", "respectauthority", 
              "nationalpride"),
  columnLabels = c("year", "ideology", "child autonomy", "imptnce of God", "resp for autho", 
              "natl pride")) +
   theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(angle = 45, size = 10),
    strip.text.y = element_text(angle = 0, hjust = 0)
    )
```

@@

This led to our first set of predictor variables, which we designate as
Recipe1: 
- Year 
- Ideology 
- Child Autonomy Index 
- Importance of God 
- Respect for Authority 
- National Pride

We also wanted to choose predictor variables that are not strongly
correlated with each other, to avoid multicollinearity. From our EDA, we
found that several predictors had correlations of 0.2 or higher with
each other. Given these correlations, we decided to make a second recipe
(`Recipe2`) with the following predictor variables, after excluding
those with a high correlation with the `godimportant` variable:

@@

@@

Note: make sure to create recipes, specify engine, make workflows, and
fit models here

@@

@@

## Final Model Output

## Discussion of Assumption, Diagnostics and Influential Points

>>>>

```{r load-data}
#| message: false
abortion_data_full<-read_csv(here::here("data/abortion-attitudes", 
                                        "wvs-usa-abortion-attitudes-data.csv"))
```

@@

Understanding public attitudes on divisive political issues is an
important way for political leaders to mobilize voters and for lawmakers
to draft laws that represent their constituents. While it is simple to
poll constituents' positions on an issue, it can be challenging to
assess the complex factors that influence and predict those stances.

Abortion is one such divisive issue in the United States, with strong
organization of pro-choice and pro-life groups in states across the
country in support of legislation for their respective sides (Ziegler,
2020). However, following the Supreme Court's 1973 decision in Roe v.
Wade, a ruling that has historically protected women's rights to
abortion without excessive government restriction in America, the issue
has risen in political salience. Both the pro-choice and pro-life
movements have gained national prominence, and the two major political
parties have polarized around the issue, with the Democratic Party in
favor of and the Republican Party against policies legalization and
increased access to abortion (Weinberger 2022). Abortion has also
increasingly become a key issue that voters consider, with an increasing
share of Americans identifying as "single-issue voters" regarding
abortion (Brenan 2020).

@@

We will conduct our analysis on data collected about Americans' abortion
attitudes, demographic information, and other ideologies, using an
EDA-informed predictive model. The data are a nationally representative
sample of the American people; hence we can infer that all the
observations are independent and the variables may show a linear
relationship. Given that this issue has become highly polarized by
political party, we predict that liberal political attitudes and youth
will correlate with belief that abortion is more justified.

@@

The response variable, `Justifiability of abortion`, is a numerical
measure on a scale of 1 to 10 on the individual person's attitude toward
whether abortion is justifiable or not. Individuals responded 1 for
"abortion is never justified" and 10 for "abortion is always justified.

@@

```{r visualization-summary}
#| warning: false
ggplot(abortion_data_full, aes(x = aj)) +
  geom_histogram(binwidth = 1, aes(y=..density..)) + 
  geom_density(color = "red") +
  labs(title = "Histogram of Abortion Attitudes",
       x = "Abortion Justifiability (1-Least, 10-Most)")

summary(abortion_data_full$aj)
```

@@

```{r eda-year}
#| warning: false
#Year of survey
ggplot(abortion_data_full, aes(x = year)) +
  geom_histogram() + 
  labs(title = "Histogram of year")
ggplot(abortion_data_full, aes(group = year, y = aj)) +
  geom_boxplot() 
```

```{r eda-ideology}
#| warning: false
# ideology
ggplot(abortion_data_full, aes(x = ideology)) +
  geom_histogram() + 
  labs(title = "Histogram of ideology")
ggplot(abortion_data_full, aes(x = ideology,group = ideology, y = aj)) +
  geom_boxplot()
```

```{r eda-cai}
#| warning: false
# Child autonomy index
ggplot(abortion_data_full, aes(x = cai)) +
  geom_histogram() + 
  labs(title = "Histogram of Child autonomy index")
ggplot(abortion_data_full, aes(x = cai, group = cai, y = aj)) +
  geom_boxplot()
```

```{r eda-god}
#| warning: false
# Importance of God
ggplot(abortion_data_full, aes(x = godimportant)) +
  geom_histogram() + 
  labs(title = "Histogram of how respondent saw God's importance")
ggplot(abortion_data_full, aes(x = godimportant, group = godimportant, 
                               y = aj)) +
  geom_boxplot()
```

```{r eda-respect}
#| warning: false
# Respect for authority
ggplot(abortion_data_full, aes(x = respectauthority)) +
  geom_histogram() + 
  labs(title = "Histogram of respect for authority")
ggplot(abortion_data_full, aes(x = respectauthority, group = respectauthority, 
                          y = aj)) +
  geom_boxplot()
```

```{r eda-pride}
#| warning: false
# National Pride
ggplot(abortion_data_full, aes(x = nationalpride)) +
  geom_histogram() + 
  labs(title = "Histogram of National pride")
ggplot(abortion_data_full, aes(x = nationalpride, group = nationalpride, y = aj)) +
  geom_boxplot()
```

@@

```{r correlations}
#| warning: false
ggpairs(abortion_data_full, 
  columns = c("year", "ideology", "cai", "godimportant", "respectauthority", 
              "nationalpride"),
  columnLabels = c("year", "ideology", "child autonomy", "imptnce of God", 
                   "resp for autho", "natl pride")) +
   theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(angle = 45, size = 10),
    strip.text.y = element_text(angle = 0, hjust = 0)
    )
```

@@

This led to our first set of predictor variables, which we designate as
Recipe1:

-   Year
-   Ideology
-   Child Autonomy Index
-   Importance of God
-   Respect for Authority
-   National Pride

We also wanted to choose predictor variables that are not strongly
correlated with each other, to avoid multi-collinearity. From our EDA,
we found that several predictors had correlations of 0.2 or higher with
each other. Given these correlations, we decided to make a second recipe
(`Recipe2`) with the following predictor variables, after excluding
those with a high correlation with the `godimportant` variable:

@@

We'll now split each of the three "recipes" into a training set (75%)
and a testing set (25%)

```{r splits}
set.seed(206)
abortion_split <- initial_split(abortion_data)
abortion_training <- training(abortion_split)
abortion_testing <- testing(abortion_split)
```

@@

Now that the data is split, we will specify the models and create
recipes for each.

```{r spec}
abortion_spec <- linear_reg() %>%
  set_engine("lm")
```

Now that our model is specified, we will create 3 recipes, each with the
corresponding predictors listed above.

The first recipe will be predicting abortion attitudes from `year`,
`ideology`, `cai`, `godimportant`, `respectauthority`, and
`nationalpride`. We will center the variable `year` so that we can have
a meaningful intercept, create dummy variables when needed, and
eliminate any zero variance predictors.

```{r recipe1}
abortion_rec1 <- recipe(aj ~ year + ideology + cai + godimportant +
                        respectauthority, nationalpride,
                        data = abortion_data) %>%
  step_center(year) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

abortion_rec1
```

Our second recipe will be predicting abortion attitudes from just `year`
and `godimportant`. We will once again center the variable `year` so
that we can have a meaningful intercept, create dummy variables when
needed, and eliminate any zero variance predictors.

```{r recipe2}
abortion_rec2 <- recipe(aj ~ year + godimportant,
                        data = abortion_data) %>%
  step_center(year) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

abortion_rec2
```

Our third and final recipe will predict abortion attitudes from `year`,
`ideology`, `cai`, and `nationalpride`. We will take the same steps from
the last 2 recipes.

```{r recipe3}
abortion_rec3 <- recipe(aj ~ year + ideology + cai + nationalpride,
                        data = abortion_data) %>%
  step_center(year) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

abortion_rec3
```

With the recipes made, we are ready to create our workflows.

```{r workflows}
abortion_wflow1 <- workflow() %>%
  add_model(abortion_spec) %>%
  add_recipe(abortion_rec1)

abortion_wflow2 <- workflow() %>%
  add_model(abortion_spec) %>%
  add_recipe(abortion_rec2)

abortion_wflow3 <- workflow() %>%
  add_model(abortion_spec) %>%
  add_recipe(abortion_rec3)
```

And finally, we will fit each of the models above

```{r fit-models}
abortion_fit1 <- abortion_wflow1 %>%
  fit(data = abortion_training)
tidy(abortion_fit1) %>%
  kable(digits = 3)

abortion_fit2 <- abortion_wflow2 %>%
  fit(data = abortion_training)
tidy(abortion_fit2) %>%
  kable(digits = 3)

abortion_fit3 <- abortion_wflow3 %>%
  fit(data = abortion_training)
tidy(abortion_fit3) %>%
  kable(digits = 3)
```

@@

Now that we have all of our models fit and displayed, we need to
interpret the results and determine the best model.

Let's start by seeing how well our models predict the abortion attitudes

```{r predict}
abortion_training_pred1 <- predict(abortion_fit1, abortion_training) %>%
  bind_cols(abortion_training %>% select(aj, year, ideology, cai, 
                                         godimportant, respectauthority, 
                                         nationalpride))
abortion_training_pred1

abortion_training_pred2 <- predict(abortion_fit2, abortion_training) %>%
  bind_cols(abortion_training %>% select(aj, year, godimportant))
abortion_training_pred2

abortion_training_pred3 <- predict(abortion_fit3, abortion_training) %>%
  bind_cols(abortion_training %>% select(aj, year, ideology, cai, 
                                         nationalpride))
abortion_training_pred3
```

A quick glance at our data might seem that the models are not that
strong at predicting the abortion attitude (aj) of a given observation.
We will look into this further now with cross-validation, then a test in
AIC and BIC statistics.

```{r cv}
set.seed(206)
folds <- vfold_cv(abortion_training, v = 10)
abortion_fit_rs1 <- abortion_wflow1 %>%
  fit_resamples(folds)

abortion_fit_rs2 <- abortion_wflow2 %>%
  fit_resamples(folds)

abortion_fit_rs3 <- abortion_wflow3 %>%
  fit_resamples(folds)
```

```{r cv-table}
cv_metrics1 <- collect_metrics(abortion_fit_rs1, summarize = FALSE) 
cv_metrics2 <- collect_metrics(abortion_fit_rs2, summarize = FALSE) 
cv_metrics3 <- collect_metrics(abortion_fit_rs3, summarize = FALSE)

cv_metrics1 %>%
  mutate(.estimate = round(.estimate, 3)) %>%
  pivot_wider(id_cols = id, names_from = .metric, values_from = .estimate) %>%
  kable(col.names = c("Fold", "RMSE", "R-squared"))

cv_metrics2 %>%
  mutate(.estimate = round(.estimate, 3)) %>%
  pivot_wider(id_cols = id, names_from = .metric, values_from = .estimate) %>%
  kable(col.names = c("Fold", "RMSE", "R-squared"))

cv_metrics3 %>%
  mutate(.estimate = round(.estimate, 3)) %>%
  pivot_wider(id_cols = id, names_from = .metric, values_from = .estimate) %>%
  kable(col.names = c("Fold", "RMSE", "R-squared"))
```

As with the predictions from before, the RMSE and R-squared values for
each of these models do not show promising results. Of the three models,
the first is seemingly the strongest model as it has the closest
R-squared values to 1. We will continue our analysis still with AIC and
BIC analysis.

```{r aic-bic}
glance(abortion_fit1) %>% 
  select(AIC, BIC)

glance(abortion_fit2) %>% 
  select(AIC, BIC)

glance(abortion_fit3) %>% 
  select(AIC, BIC)
```

From the above AIC and BIC values, we have further evidence that model 1
is the strongest model of the three. Model 1 has the lowest AIC and BIC
values, which are usually penalized for having more predictors, but in
this case the model is strong enough to overcome that penalty.

Before we discuss results, we will take a quick look at the resulting
confidence intervals from our fitted models and the resulting
coefficients.

```{r ci}
tidy(abortion_fit1, conf.int = TRUE) %>%
  kable(digits = 3)
tidy(abortion_fit2, conf.int = TRUE) %>%
  kable(digits = 3)
tidy(abortion_fit3, conf.int = TRUE) %>%
  kable(digits = 3)
```

@@

From the analysis of the models through multiple different calculations,
we know that of our three models, model 1 is the strongest, predicting
abortion attitude from `year`, `ideology`, `cai`, `godimportant`,
`respectauthority`, and `nationalpride`. Given that model 1 is our
strongest model, it must also be noted that all 3 of our models had
difficulty predicting an accurate attitude towards abortion of a given
observation. This is most likely due to the data having a large number
of very different observations given the same `aj` score. In model 1,
however, our p-values for the coefficients given are all at an
acceptable range except for that of `nationalpride`. This means that
every other predictor was able to be helpful in predicting the response
variable.

## Discussion + Conclusion

Ultimately, our investigation in the correlation between attitudes on
the justifiability of abortion and demographic factors + personal
attitudes towards other issues presented some problems. Through our
model creation and comparison, we are able to determine that the best
combination of predictors for abortion attitudes are `year`, `ideology`,
`cai`, `godimportant`, `respectauthority`, and `nationalpride`. While
every one of these predictors are shown to be helpful in predicting the
response variable, the model itself is generally weak although it is the
best of our 3.

Unfortunately, given our data set, we are limited by the truthfulness
and accuracy of each observation. While we can assume each observation
is answered honestly and thoughtfully, there was the option to leave
some answers empty, and some predictors are quite hard to place nicely
into a 1-10 scale (take ideology for example). Steps that could be taken
to counter this problem are to either categorize some of these 1-10
scales into even smaller categories (like 3/4) to possibly show better
trends, or even more categories (like 1-50 scale) to increase accuracy
of responses and also possibly reveal more trends.

With abortion being such a divisive issue in our society, it is
incredibly difficult to pinpoint just a few predictors to try and
accurately tell how any one person might feel about abortion. Knowing
that, however, we do not expect that abortion attitudes are impossible
to more accurately predict. We have shown through our modeling and
analysis that some of the predictors presented are helpful in predicting
our response variable. Given this, it is not unreasonable to assume that
there might be even more helpful predictors to use in place, or in
addition to the ones we have selected. For future work, we might be able
to produce stronger models given additional demographic/personality
predictors to work with.

### 2022-06-21 written-report.qmd
FINAL REORG

plot_grid(title_factor, top_row, title_numeric, bottom_row, caption3, 
          ncol = 1, rel_heights = c(0.1, 1,0.1,1,0.1))

@@

plot_grid(top_row1, c3, caption4, caption5, caption6,
          ncol = 1, rel_heights = c(1, 1,0.1,0.1,0.1))

@@

Despite the model confirming our initial research question, we have reason to question the quality of the model, and whether the model accurately fits the data on abortion attitudes. The first reason for questioning our model’s quality is the low R-squared value, which indicates that the variance of our predictors may not greatly account for the variance in abortion attitudes. The second reason for questioning our model’s quality is that our model does not fit the linearity and constant variance conditions. Upon closer examination of the data, we can see that the distribution of our response variable was trimodal, with peaks at the extremes and in the middle. This likely led to the data having a large number of very different observations that had different predictor values, but the same `aj` score, and may be one explanation our model’s weaknesses.

Given the nature of public opinion on political issues, we are likely limited by the precision and accuracy of each observation in our dataset. Previous political science research on public opinion indicates that Americans’ political beliefs are inconsistent and volatile. For example, most Americans’ political beliefs have been found to be shaped by partisan and group loyalties, rather than ideology or any specific set of issue positions. Indeed, even for committed partisans, one individual’s issue positions are likely to to be scattered across the ideological spectrum, and may even change over short periods of time (Achen and Bartels, 2017). This inconsistency between partisanship, ideology, and issue positions may account for the non-normal distribution of abortion attitudes in the dataset and the model’s poor fit to the data. In addition, although the data are supposed to capture a nationally representative sample of the US population, we do not know if they are weighted by other factors that may represent important distinctions in the US population. This possibility of non-weighting may have influenced our model’s quality as well.

Future work may build on this analysis and correct some of its weaknesses. First, the data should be weighted by more relevant factors to ensure that the sample is as representative as possible of the US population. Relevant factors may include, but are not limited to, partisan affiliation, race, religion, socioeconomic status, and geographic area. In addition, it may be helpful to consider other predictors that may influence abortion attitudes, such as partisan affiliation and race. However, it is important to keep in mind that abortion is an incredibly divisive and complex issue, so it is not an easy task to precisely pinpoint any individual’s abortion attitude, no matter which predictors are used.

@@

Miller, Steven V. (2016, Jul 31).  “Attitudes on the Justifiability of Abortion in the United States.”  GitHub.  https://github.com/svmiller/wvs-usa-abortion-attitudes

>>>>

plot_grid(title_factor, top_row, title_numeric, bottom_row, caption3, 
          ncol = 1, rel_heights = c(0.1, 1,0.1,1,0.2))

@@

plot_grid(top_row1, c3, caption4, caption5, caption6,
          ncol = 1, rel_heights = c(1, 1,0.2,0.2,0.2))

@@

Despite the model confirming our initial research question, we have
reason to question the quality of the model, and whether the model
accurately fits the data on abortion attitudes. The first reason for
questioning our model's quality is the low R-squared value, which
indicates that the variance of our predictors may not greatly account
for the variance in abortion attitudes. The second reason for
questioning our model's quality is that our model does not fit the
linearity and constant variance conditions. Upon closer examination of
the data, we can see that the distribution of our response variable was
trimodal, with peaks at the extremes and in the middle. This likely led
to the data having a large number of very different observations that
had different predictor values, but the same `aj` score, and may be one
explanation our model's weaknesses.

Given the nature of public opinion on political issues, we are likely
limited by the precision and accuracy of each observation in our
dataset. Previous political science research on public opinion indicates
that Americans' political beliefs are inconsistent and volatile. For
example, most Americans' political beliefs have been found to be shaped
by partisan and group loyalties, rather than ideology or any specific
set of issue positions. Indeed, even for committed partisans, one
individual's issue positions are likely to to be scattered across the
ideological spectrum, and may even change over short periods of time
(Achen and Bartels, 2017). This inconsistency between partisanship,
ideology, and issue positions may account for the non-normal
distribution of abortion attitudes in the dataset and the model's poor
fit to the data. In addition, although the data are supposed to capture
a nationally representative sample of the US population, we do not know
if they are weighted by other factors that may represent important
distinctions in the US population. This possibility of non-weighting may
have influenced our model's quality as well.

Future work may build on this analysis and correct some of its
weaknesses. First, the data should be weighted by more relevant factors
to ensure that the sample is as representative as possible of the US
population. Relevant factors may include, but are not limited to,
partisan affiliation, race, religion, socioeconomic status, and
geographic area. In addition, it may be helpful to consider other
predictors that may influence abortion attitudes, such as partisan
affiliation and race. However, it is important to keep in mind that
abortion is an incredibly divisive and complex issue, so it is not an
easy task to precisely pinpoint any individual's abortion attitude, no
matter which predictors are used.

@@

Miller, Steven V. (2016, Jul 31). "Attitudes on the Justifiability of
Abortion in the United States." GitHub.
https://github.com/svmiller/wvs-usa-abortion-attitudes
